---
title: "STA243 Hw2 - Eric Kalosa-Kenyon"
output:
  html_document:
    toc: yes
---

This assignment contains solutions to the genetic programming and simulated
annealing problems. Here both algorithms are implemented in R and used to solve
a piecewise constant regression problem and a small traveling salesman problem
respectively.

# Simulated annealing

## Setup
```{r, message=F, warning=F}
### Purpose of this algorithm
#   The goal of this program is to find an optimal solution for the traveling
#   salesman problem. The method used is simulated annealing.

### Pseudocode overview
# Generate starting sample s
# For k in k_0 : k_max
#   T = temp(k_0/k_max)
#   sample from neighborhood N(s) -> s'
#   if H(s')<H(s):
#       s <- s'
#   else:
#       if e^(-(H(s')-H(s))/T) > runif(0,1):
#           s <- s'
# return s
# Plot result

##### BEGIN: Code
rm(list=ls())

## Imports
library("ggplot2")
library("dplyr")

## Define distance matrix
raw = c(
    0, 1, 2, 4, 9, 8, 3, 2, 1, 5, 7, 1, 2, 9, 3,
    1, 0, 5, 3, 7, 2, 5, 1, 3, 4, 6, 6, 6, 1, 9,
    2, 5, 0, 6, 1, 4, 7, 7, 1, 6, 5, 9, 1, 3, 4,
    4, 3, 6, 0, 5, 2, 1, 6, 5, 4, 2, 1, 2, 1, 3,
    9, 7, 1, 5, 0, 9, 1, 1, 2, 1, 3, 6, 8, 2, 5,
    8, 2, 4, 2, 9, 0, 3, 5, 4, 7, 8, 3, 1, 2, 5,
    3, 5, 7, 1, 1, 3, 0, 2, 6, 1, 7, 9, 5, 1, 4,
    2, 1, 7, 6, 1, 5, 2, 0, 9, 4, 2, 1, 1, 7, 8,
    1, 3, 1, 5, 2, 4, 6, 9, 0, 3, 3, 5, 1, 6, 4,
    5, 4, 6, 4, 1, 7, 1, 4, 3, 0, 9, 1, 8, 5, 2,
    7, 6, 5, 2, 3, 8, 7, 2, 3, 9, 0, 2, 1, 8, 1,
    1, 6, 9, 1, 6, 3, 9, 1, 5, 1, 2, 0, 5, 4, 3,
    2, 6, 1, 2, 8, 1, 5, 1, 1, 8, 1, 5, 0, 9, 6,
    9, 1, 3, 1, 2, 2, 1, 7, 6, 5, 8, 4, 9, 0, 7,
    3, 9, 4, 3, 5, 5, 4, 8, 4, 2, 1, 3, 6, 7, 0
    )
D = matrix(raw, sqrt(length(raw)), sqrt(length(raw)))

## Parameterize script
N = dim(D)[1] # number of cities
M = 15 # number of times to anneal
k_max = 1000 # steps in the annealing process
```

## Subroutines

```{r}
## Subroutine definitions
temp = function(r){
    ## Temperature schedule
    stopifnot(r > 0 || r <= 1)

    # when r is large, temperature should be low
    # temp should always be positive, real
    tt = -log(r)
    return(tt)
}

path_len = function(s, DM=D){
    r = 0
    N = length(s)
    for(i in 1:(N-1)){ # travel
        r = r + DM[s[i],s[i+1]]
    }
    r = r + DM[s[N],s[1]] # go home
    return(r)
}

nhbd = function(s, k = 2){
    # Generate a neighborhood around path (s = e.g. [1,4,2,3] for N=4)
    #   which consists of all sigma(s) where sigma is a k-cycle rotation with k
    #   non-zero elements.

    # Check input
    N = length(s)
    stopifnot(N > 2)
    stopifnot(k >= 2) # 1 cycle neighborhood is identity
    stopifnot(k <= N) # there are no N+1 cycles in Z/Z_{N}

    # Generate k-cycle
    K = sample(1:N, k)

    # Rotate s by K
    s2 = s
    for(i in 1:(k-1)){
        s2[K[i]] = s[K[i+1]]
    }
    s2[K[k]] = s[K[1]]

    # return result
    return(s2)
}
```

## Unit tests
```{r}
## Tests
test_all_subroutines = function(){
    ## Unit tests

    # Test neighborhood generator
    for(i in 1:100){
        k = floor(runif(1)*18) + 2
        N = k + floor(runif(1)*10) + 1
        s = sample(1:N)
        n = nhbd(s, k)
        stopifnot(sum(n != s) == k)
    }
    print("nhbd(s,k) passed!")

}
```

## Simulate
```{r}
## Anneal
results = data.frame(
    Step=integer(),
    Length=integer(),
    M=integer(),
    stringsAsFactors=F
)

# anneal M times to avoid local minima
best_paths = list()
for(m in 1:M){

    s0 = sample(1:N) # Starting path sampled uniformly at random

    for(k in 1:k_max){

        tt = temp(k/k_max) # Get temperature from temperature schedule
        s1 = nhbd(s0) # Sample uniformly at random from the neighborhood of s0
        h0 = path_len(s0) # Calculate energy of object
        h1 = path_len(s1)

        if(h1 < h0){ # If the new path is shorter
            s0 = s1 # take it
        }else{
            R = runif(1)
            P = exp(-(h1-h0)/tt)

            if(P>R){ # Metropolis-Hastings
                s0 = s1
            }
        }

        # record results
        results[dim(results)[1]+1,] = c(k, path_len(s0), m)

    }

    best_paths[[length(best_paths)+1]] = s0
}
results = results[results$Step!=k_max,]
best_path =
    best_paths[[
        results$M[
            tail(which(
                results$Length==min(results$Length)
            ),1)
        ]
    ]]
```

## Results
```{r}
## Plot result
plt_ann = ggplot(results, aes(x=Step, y=Length, color=factor(M))) +
    geom_line() +
    ggtitle("Annealing the Traveling Salesman problem") +
    xlab("Step") + ylab("Path length") +
    labs(color="Annealing\nrun")
plt_ann
```

The minimal path length calculated here is `r min(results$Length)` using the
path `r best_path`.

# Genetic algorithm

## Setup
```{r, message=F, warning=F}
### Pseudocode overview
# Generate observations used for regression
# Parameterize algorithm
# Initialize random population
# Until convergence
#   Calculate fitness in current generation
#   Mate individuals proportional to fitness
# Plot result

##### BEGIN: Code
rm(list=ls())

### Imports
library("ggplot2")
library("dplyr")

### Parameterize script
K = 100  # Number of chromosomes in each generation
Gmax = 30  # Maximum number of generations
Pcross = 0.9 # Crossover rate
Pmutate = 0.05 # Mutation rate
which_penalty = "AIC" # "AIC" or "MDL"
```

## Subroutines
```{r}
### BEGIN: Subroutines
random_chromosome = function(num_genes, nucleobases=c(0,1), max_non0=25){
    # Generates a chromosome
    #   lenght of chromosome is (num_genes)
    #   each element of the chromosome is an element of (nucleobases)
    non0 = sample(1:num_genes, max_non0)
    x = rep(nucleobases[1], num_genes)
    x[non0] = sample(nucleobases, max_non0, replace=TRUE)
    x = paste(x, collapse='')
    return(x)
}

mate_chromosomes = function(parents, xr=0.1, mr=0.1){
    # Produce a single offspring from parent chromosomes (p1, p2) with a
    # crossover rate (xr) and a mutation rate (mr)

    p1 = parents[1]
    p2 = parents[2]
    stopifnot(nchar(p1) == nchar(p2))

    child = ""
    w = 1 # crossover parameter
    for(i in 1:nchar(p1)){

        # Give child parents' genes for each gene (i)
        #   from p1 if w == 1
        #   from p2 if w == 0
        if(w){
            chr = substr(p1, i, i) # get i'th character from parent 1
        }else{
            chr = substr(p2, i, i) # get i'th character from parent 2
        }

        # Mutate gene
        if(runif(1)<mr){
            chr = as.character((as.integer(chr) + 1) %% 2)
        }

        # Extend child with the i'th parental character
        child = paste(child, chr, sep="")

        # Crossover if required
        if(runif(1)<xr){
            w = (w + 1) %% 2
        }
    }
    return(child)
}

regression_loss = function(reg, raw){
    # Sum of square error betweeen regression and raw
    # len(reg) == len(raw)
    return(sum((reg-raw)^2))
}

regress_chromosome = function(chrom, raw){
    # return c(fhat(x))
    # chrom is a string of "0001110101.."
    # raw is a c(num), len(chrom) == len(raw) - 1

    # Initialize working objects
    segment_observations = c(raw[1])
    regression = c()

    for(i in 1:nchar(chrom)){ # for each gene in the chromosome
        g = as.integer(substr(chrom, i, i)) # extract the gene from the string

        if(g){ # if the gene indicates a breakpoint
            # calculate the regression for that segment

            regression = c(regression,
                           rep(mean(segment_observations),
                               length(segment_observations)
                               )
                           )
            segment_observations = c(raw[i+1]) # reset to next segment

        }else{ # if the gene does not indicate a breakpoint
            # extend the segment
            segment_observations = c(segment_observations, raw[i+1])
        }
    }

    # calculate final segment regression
    regression = c(regression,
                   rep(mean(segment_observations),
                       length(segment_observations)
                       )
                   )

    return(regression)
}

calc_AIC = function(n, k, loss){
    # AIC is k-ln(sum sq losses)
    #   n is positive integer
    #   k is num params
    #   loss is losses
    #   k and loss can be c() as long as they have the same length
    # Note that this AIC can be alternatively parameterized

    r = 2*log(n)*k + n*log(loss/n)
    return(r)
}

count_params = function(vec_of_chromosomes){
    # given a vector of chromosomes (each is "00111010111...")
    #   count the number of segments this represents i.e. the number of
    #   parameters that are implicitly contained in this chromosome.

    r = sapply( # vector of integers summing number of breakpoints
            lapply(
                strsplit(vec_of_chromosomes, ""),
                as.integer),
            sum) + 1 # number of constants is breaks + 1
    return(r)
}

fitness = function(population, raw){
    # Calculate fitness for each individual in the population relative to the
    #   raw data.
    #   Complexity is either "AIC" or "MDL". Raw is the raw data to fit.
    # Return a vector of floats between 0 and 1.
    #   The AIC and MDL are normalized to [0,1]

    # Check that input is ok
    stopifnot(nchar(population$Chromosomes[1]) + 1 == length(raw))

    # Initialize parameters
    losses = c()
    K = dim(population)[1]  # (K) is number of individuals in population
    n = length(raw) # (n) is the number of data observations

    # Calculate losses for each individual in the generation
    for(k in 1:K){
        x = population$Chromosome[k] # extract a chromosome
        regr = regress_chromosome(x, raw) # calculate the regression
        loss = regression_loss(regr, raw) # and calculate the loss
        losses = c(losses, loss) # record that individual's loss
    }

    ## Penalize for complexity
    ns_params = count_params(population$Chromosome)
    AIC = calc_AIC(n, ns_params, losses)

    # calculate sum of log(\hat{n_j}) for MDL calculation
    # sum_n_hat = for each chromosome, length of 0s between 1s plus 1
    # TODO: refactor into a separate subroutine
    #   MDL = calc_MDL(n, ns_params, losses)
    sum_n_hat = sapply(
                    lapply( # for each chromosome, get \hat{n_j}'s
                        strsplit( # get strs of 0's (whose lengths = n_j-1)
                            population$Chromosome,
                            "1"
                        ),
                        nchar),
                    (function (x) sum(log(x+1))) # sum(log(\hat{n_j}))
                )
    MDL = ns_params*log(n) +
        1/2*sum_n_hat +
        n/2*log(losses/n)

    return(list(AIC, MDL))
}

plot_chromosome = function(chrom, raw_df, color="blue"){
    # return a ggplot object with piecewise constants plotted on the scatter

    fhat = regress_chromosome(chrom, raw_df$y)
    fhat_df = data.frame(x=raw_df$x, y=fhat)

    plt = ggplot(raw_df, aes(x=x, y=y)) +
        geom_point() +
        geom_line(data=fhat_df, color=color) +
        xlab("Index") + ylab("Value")

    return(plt)
}

plot_generation = function(gen, raw_df, wh_score="AIC"){
    # return a ggplot object with piecewise constants plotted on the scatter
    #   for each chromosome in the generation
    # Currently, colorized lines based on AIC/MDL are unimplemented

    wh_gen = gen$Generation[1]

    # prepare base plot
    plt = ggplot(raw_df, aes(x=x, y=y)) +
        geom_point() +
        xlab("Index") + ylab("Value") +
        ggtitle(paste("Generation", wh_gen, wh_score))

    # prepare regression data for plotting
    fhat_df = data.frame(matrix(ncol = 5, nrow = 0))
    colnames(fhat_df) = c("x","y","Group","AIC","MDL")
    for(i in 1:dim(gen)[1]){
        chrom = gen$Chromosome[i]
        fhat = regress_chromosome(chrom, raw_df$y)
        fhat_df_i = data.frame(x=raw_df$x, y=fhat)
        fhat_df_i$Group = i
        fhat_df_i$AIC = as.numeric(gen$AIC[i])
        fhat_df_i$MDL = as.numeric(gen$MDL[i])
        fhat_df = rbind(fhat_df, fhat_df_i)
    }

    # add geomlines for aic or mdl depending on which was passed into this fxn
    if(wh_score=="AIC"){
        plt = plt + geom_line(data=fhat_df, aes(color=AIC, group=Group))
    }else if(wh_score=="MDL"){
        plt = plt + geom_line(data=fhat_df, aes(color=MDL, group=Group))
    }

    # plot AIC or MDL in a color gradient
    plt = plt + scale_colour_gradientn(colours=rainbow(4))

    return(plt)
}

best_score_and_org = function(population, wh_score){
    # Determine best score in the population and which organism achieved it
    if(wh_score=="AIC"){
        best_score = min(as.numeric(population$AIC))
        which_organism_is_best = which(
            as.numeric(population$AIC) == best_score)[1]
    }else if(wh_score=="MDL"){
        best_score = min(as.numeric(population$MDL))
        which_organism_is_best = which(
            as.numeric(population$MDL) == best_score)[1]
    }

    return(list(best_score, which_organism_is_best))
}

### END: Subroutines
```

## Simulate

```{r}
### BEGIN: Simulation

## Generate observations

# True data given in homework
truefunction<-function(x){
    t <- c(0.1, 0.13, 0.15, 0.23, 0.25, 0.4, 0.44, 0.65, 0.76, 0.78, 0.81)
    h <- c(4, -5, 3, -4, 5, -4.2, 2.1, 4.3, -3.1, 2.1, -4.2)
    temp <- 0
    for(i in 1:11) {
        temp <- temp + h[i]/2 * (1 + sign(x - t[i]))
    }
    return(temp)
}
ndata = 512
xdata = (0:(ndata-1))/ndata
ydata = truefunction(xdata)
set.seed(0401)
N = length(ydata)
raw = ydata + rnorm(N)/3

## Initialize major objects
pop_col_names = c("Chromosome", "AIC", "MDL", "Generation")
pop = data.frame(matrix(ncol = 4, nrow = 0))   # Holds all individuals
colnames(pop) = pop_col_names

## Initialize first generation
g = 1
for(i in 1:K){
    x = random_chromosome(num_genes=N-1) # returns string of 1s and 0s
    individual = c(x, NA, NA, g)
    pop[dim(pop)[1]+1,] = individual
}

## Calculate fitness and mate each generation
for(g in 2:Gmax){
    parent_generation = pop[pop$Generation == g-1,]
    parent_aic_mdl = fitness(parent_generation, raw)
    # Put parent fitness into main dataframe
    pop[rownames(parent_generation),]$AIC = parent_aic_mdl[[1]]
    pop[rownames(parent_generation),]$MDL = parent_aic_mdl[[2]]

    # Update current working object
    parent_generation$AIC = parent_aic_mdl[[1]]
    parent_generation$MDL = parent_aic_mdl[[2]]

    # invert AIC and MDL for sampling
    parent_generation$nAIC =
        -(parent_generation$AIC - min(parent_generation$AIC))
    parent_generation$nAIC =
        parent_generation$nAIC - min(parent_generation$nAIC) + 0.001
    parent_generation$nMDL =
        -(parent_generation$MDL - min(parent_generation$MDL))
    parent_generation$nMDL =
        parent_generation$nMDL - min(parent_generation$nMDL) + 0.001

    for(k in 1:K){
        # Sample parents according to AIC or MDL
        if(which_penalty == "AIC"){
            parents = sample_n(parent_generation,
                             2,
                             replace=TRUE, # do not require sexual reproduction
                             weight=nAIC)
        } else if(which_penalty == "MDL"){
            parents = sample_n(parent_generation,
                             2,
                             replace=TRUE,
                             weight=nMDL)
        }

        parent_chromosomes = parents$Chromosome
        child_chromosome = mate_chromosomes(
                                parent_chromosomes,
                                Pcross,
                                Pmutate
                            )
        child = c(child_chromosome, NA, NA, g)
        pop[dim(pop)[1]+1,] = child
    }

    # TODO: add score checking convergence criterion

}

# Calculate final AIC/MDL
last_gen = pop[pop$Generation == Gmax,]
aic_mdl = fitness(last_gen, raw)
# Put last fitness into main dataframe
pop[rownames(last_gen),]$AIC = aic_mdl[[1]]
pop[rownames(last_gen),]$MDL = aic_mdl[[2]]

### END: Simulation
```

## Results and discussion
```{r}
### BEGIN: Plotting

# Plot observations
raw_df = data.frame(y=raw, x=1:length(raw))
plt_raw = ggplot(raw_df, aes(x=x, y=y)) +
    geom_point() +
    xlab("Index") + ylab("Value") +
    ggtitle("Raw observations")
plt_raw

# Plot first and last generation colored by AIC or MDL
fgen = pop[pop$Generation==1,]
lgen = pop[pop$Generation==Gmax,]
plt_fgen = plot_generation(fgen, raw_df, which_penalty)
plt_lgen = plot_generation(lgen, raw_df, which_penalty)
plt_fgen
plt_lgen

# Plot best AIC or MDL
if(which_penalty=="AIC"){
    best_score = min(as.numeric(pop$AIC))
    which_organism_is_best = which(as.numeric(pop$AIC) == best_score)
}else if(which_penalty=="MDL"){
    best_score = min(as.numeric(pop$MDL))
    which_organism_is_best = which(as.numeric(pop$MDL) == best_score)
}

best_chromosome = pop$Chromosome[which_organism_is_best]
plt_best = plot_chromosome(best_chromosome, raw_df, color="green") +
    ggtitle(
        paste(
            "Chromosome ", substr(best_chromosome, 1, 12), "... ",
            "with ", which_penalty, "=", best_score,
            sep=""
        ))
plt_best

### END: Plotting

##### END: Code
```

For the most part, this problem is an engineering exercise. We have studied
little about the statistical properties of genetic methods, so there is little
to report for semantic learning - most of my learning in this problem is
skill-based. I got better at programming in R. I'm grateful for the
opportunity to explore this methodology directly. The method is undeniably
functional in this context at least, even if we don't know much about its
statistical properties.

One trick I learned was that if I observe a low-p environment, I should
instantiate my chromosomes with low-p. This strong prior on the number of
parameters is easy enough to replicate in the fitness function with e.g. setting
fitness of all chromosomes having more than p parameters to 0, but this is
simply a matter of taste. My solution to incorporating this strong prior can be
observed in the initialization step that includes "max_non0s=20", restricting
the number of breakpoints in each chromosome in the initial population.
